Generating Random Logic Programs Using Constraint Programming

### 2019-12-28 paper/paper.tex
Refactoring + one more anti-symmetry constraint

\newcommand{\variable}[1]{\texttt{\textup{#1}}}

\author{Paulius Dilkas}
\title{Using Constraint Programming to Generate Random Logic Programs}

@@

Motivation:
\begin{itemize}
\item Generating random programs that generate random data.
  %For example, enforcing that the decision is independent of gender.
\item Learning: how this can be used for (targeted) learning, when (atomic)
  probabilities can be assigned based on counting and we can have extra
  constraints. A more primitive angle: generate structures, learn weights.
\item Describe how the trees work with many examples.
\item Have a 'Constraint' environment.
\end{itemize}

@@

\begin{itemize}
\item Support for constants and multiple variables.
  \begin{itemize}
  \item Each clause is defined for each predicate with standard variable names
    (X, Y, etc.).
  \item There is a list of constants and a list of predicates with their
    arities. Maybe also a list of variables.
  \end{itemize}
\item Make negative cycle detection use the graph representation.
\item Finish the propagation algorithm for conditional independence. The
  propagation algorithm for independence is extended with masks that are either
  potential or definite. Masking happens in two stages: first, we mask
  expressions within formulas, and then predicates. Masking algorithm uses an
  algorithm for perfect bipartite matching.
\item Show that the set of all ProbLog programs is equal to the set of programs
  I can generate (alternatively, show that, given any ProbLog program, there are
  parameter values high enough to generate it).
\item Given fixed parameters, use combinatorial arguments to calculate how many
  different programs there are and check that I'm generating the same number.
\item Formal definition (here and in the predicate invention paper): two
  predicates are independent if all of their groundings are independent.
\item Describe: entailment checking for cycles and negative cycles.
\end{itemize}

@@

Parameters:
\begin{itemize}
\item maximum number of solutions
\item \variable{maxNumNodes} (in the tree representation of a clause)
\item list of predicates with their variables
\item maximum number of clauses
\item option to forbid all cycles or just negative cycles
\item list of probabilities that are randomly assigned to clauses: $\{ 0.1,
  0.2, \dots, 0.9, 1, 1, 1, 1, 1, 1 \}$
\end{itemize}

Decision variables: % TODO: rename these things
% TODO: specify the decision variables inside clauses and their heads
\begin{itemize}
\item \variable{IntVar[] clauseAssignments}: a predicate or disabled
\item \variable{Clause[] clauses}
\item \variable{Head[] clauseHeads}
\end{itemize}

@@

\begin{constraint}[Each predicate gets at least one clause]
  \variable{numDisabledClauses} is defined by a \variable{count} constraint.
  \[
    \variable{numDistinctValues} =
    \begin{cases}
      \variable{numPredicates} + 1 & \text{if } \variable{numDisabledValues} > 0 \\
      \variable{numPredicates} & \text{otherwise}
    \end{cases}
  \]
  (also constrained using the \variable{nValues} constraint).
\end{constraint}

\begin{constraint}[\variable{clauseAssignments} are sorted]
  \[
    \variable{clauseAssignments}[i-1] = \variable{clauseAssignments}[i] \implies
    \variable{clause}[i-1] \preceq \variable{clause}[i].
  \]
\end{constraint}

@@

\begin{definition}
  The body of a clause is defined by:
  \begin{itemize}
  \item \variable{IntVar[] treeStructure}
    \begin{itemize}
    \item $\texttt{treeStructure}[i] = i$: the $i$-th node is a root.
    \item $\texttt{treeStructure}[i] = j$: the $i$-th node's parent is node $j$.
    \end{itemize}
  \item \variable{IntVar[] treeValues}: $\neg$, $\land$, $\lor$, $\top$, and any
    predefined predicates with variables.
  \end{itemize}
\end{definition}
% TODO: examples
% TODO: update the info on treeValues

Auxiliary variables: $\variable{numNodes}, \variable{numTrees} \in \{ 1, \dots,
\variable{maxNumNodes} \}$.

@@

\begin{constraint}
  $\variable{numTrees} + \variable{numNodes} = \variable{maxNumNodes} + 1$.
\end{constraint}

@@

\begin{constraint}
  For $i = 0, \dots, \variable{maxNumNodes} - 1$, if $\variable{numNodes} \le
  i$, then
  \[
    \variable{treeStructure}[i] = i \quad \text{and} \quad
    \variable{treeValues}[i] = \top,
  \]
  else
  \[
    \variable{treeStructure}[i] < \variable{numNodes}.
  \]
\end{constraint}

\begin{constraint}
  For $i = 0, \dots, \variable{maxNumNodes} - 1$,
  \begin{itemize}
  \item has 0 children $\iff$ $\variable{treeValues}[i]$ is a predicate;
  \item has 1 child $\iff$ $\variable{treeValues}[i] = \neg$;
  \item has $>1$ child $\iff$ $\variable{treeValues}[i] \in \{ \land, \lor \}$.
  \end{itemize}
\end{constraint}

\begin{constraint}
  For $i = 0, \dots, \variable{maxNumNodes} - 1$,
  \[
    \variable{treeStructure}[i] \ne i \implies \variable{treeValues}[i] \ne
    \top.
  \]
\end{constraint}

If the clause should be disabled, $\variable{numNodes} = 1$ and
$\variable{treeValues}[0] = \top$.

\begin{constraint}
  Adjacency matrix representation:
  \[
    A[i][j] = 0 \iff \nexists k: \variable{clauseAssignments}[k] = j \text{ and }
    i \in \variable{clauses}[k].\variable{treeValues}
  \]
\end{constraint}

@@

>>>>

\newcommand{\variable}[1]{\texttt{\textup{#1}}}
\newcommand{\predicates}{\mathcal{P}}
\newcommand{\variables}{\mathcal{V}}
\newcommand{\constants}{\mathcal{C}}
\newcommand{\tokens}{\mathcal{T}}
\newcommand{\arities}{\mathcal{A}}
\newcommand{\maxArity}{\textsc{max\_arity}}
\newcommand{\maxNumNodes}{\textsc{max\_num\_nodes}}
\newcommand{\maxNumClauses}{\textsc{max\_num\_clauses}}

\author{Paulius Dilkas}
\title{Generating Random Logic Programs Using Constraint Programming}

@@

Motivation:
\begin{itemize}
\item Generating random programs that generate random data.
  %For example, enforcing that the decision is independent of gender.
\item Learning: how this can be used for (targeted) learning, when (atomic)
  probabilities can be assigned based on counting and we can have extra
  constraints. A more primitive angle: generate structures, learn weights.
\end{itemize}

We will often use $\Box$ as a special domain value to indicate some kind of
exception.

@@

\begin{itemize}
\item Make negative cycle detection use the graph representation.
\item Show that the set of all ProbLog programs is equal to the set of programs
  I can generate (alternatively, show that, given any ProbLog program, there are
  parameter values high enough to generate it).
\item Given fixed parameters, use combinatorial arguments to calculate how many
  different programs there are and check that I'm generating the same number.
\item Formal definition (here and in the predicate invention paper): two
  predicates are independent if all of their groundings are independent.
\item Describe: entailment checking for cycles and negative cycles.
\item A constraint for logical equivalence.
\item Rename everything into something more appropriate.
\end{itemize}

@@

Parameters:
\begin{itemize}
\item maximum number of solutions
\item \maxNumNodes{} (in the tree representation of a clause)
\item \maxNumClauses{}
\item option to forbid all cycles or just negative cycles
\item a list of probabilities that are randomly assigned to clauses: $\{ 0.1,
  0.2, \dots, 0.9, 1, 1, 1, 1, 1, 1 \}$.
\item a list of predicates $\predicates{}$,
\item a list of their arities $\arities{}$,
  \begin{itemize}
  \item $\maxArity{} = \max \arities{}$.
  \end{itemize}
\item a list of variables $\variables{}$,
\item and a list of constants $\constants{}$.
\end{itemize}

We also define $\tokens{} = \{ \neg, \land, \lor, \top \}$.

Decision variables:
% TODO: specify the decision variables inside clauses and their heads
\begin{itemize}
\item \variable{Body[] bodiesOfClauses}
\item \variable{Head[] headsOfClauses}: a list of length \maxNumClauses{}
\end{itemize}

@@

\begin{constraint}[Each predicate gets at least one clause]
  Let $P = \{ h.\variable{predicate} \mid h \in \variable{clauseHeads} \}$. Then
  \[
    \variable{nValues}(P) =
    \begin{cases}
      \variable{numPredicates} + 1 & \text{if } \variable{count}(\Box, P) > 0 \\
      \variable{numPredicates} & \text{otherwise.}
    \end{cases}
  \]
\end{constraint}

\begin{constraint}
  Let $\prec$ be any total order defined over bodies of clauses, and let
  $\preceq$ be its extension with equality (in the same way as $\le$ extends
  $<$).
  \[
    \variable{headsOfClauses}[i-1].\variable{predicate} =
    \variable{headsOfClauses}[i].\variable{predicate} \implies
    \variable{bodiesOfClauses}[i-1] \preceq \variable{bodiesOfClauses}[i].
  \]
\end{constraint}
For example, $\preceq$ can be implemented as \variable{lexLessEq} over the
decision variables of each body.
% TODO: might need to include clause heads as well.

\section{Atoms}

\begin{definition}
  An \emph{atom} is a $\variable{predicate} \in \tokens{} \cup \predicates{}$ and a
  list of \variable{arguments} of length \maxArity{} in $\variables{} \cup
  \constants{} \cup \{ \Box \}$, where $\Box$ means the position is either
  reserved for a variable, or disabled. The atom's \variable{arity} is a number
  in $[0, \maxArity{}]$ defined by a \variable{table} constraint,
  according to the \variable{predicate}.
\end{definition}

\begin{constraint}
  For $i = 0, \dots, \maxArity{} - 1$,
  \[
    i \ge \variable{arity} \implies \variable{arguments}[i] = 0.
  \]
\end{constraint}

@@

\begin{definition}
  The body of a clause is defined by:
  \begin{itemize}
  \item \variable{treeStructure}: list of length \maxNumNodes{} with
    domain $[0, \maxNumNodes{}]$.
    \begin{itemize}
    \item $\variable{treeStructure}[i] = i$: the $i$-th node is a root.
    \item $\variable{treeStructure}[i] = j$: the $i$-th node's parent is node $j$.
    \end{itemize}
  \item \variable{treeValues}: \maxNumNodes{} atoms.
  \end{itemize}
\end{definition}
% TODO: examples

Auxiliary variables: $\variable{numNodes}, \variable{numTrees} \in \{ 1, \dots,
\maxNumNodes{} \}$.

@@

\begin{constraint}
  $\variable{numTrees} + \variable{numNodes} = \maxNumNodes{} + 1$.
\end{constraint}

@@

\begin{constraint}
  For $i = 0, \dots, \maxNumNodes{} - 1$, if $\variable{numNodes} \le
  i$, then
  \[
    \variable{treeStructure}[i] = i \quad \text{and} \quad
    \variable{treeValues}[i].\variable{predicate} = \top,
  \]
  else
  \[
    \variable{treeStructure}[i] < \variable{numNodes}.
  \]
\end{constraint}

\begin{constraint}
  For $i = 0, \dots, \maxNumNodes{} - 1$,
  \begin{itemize}
  \item has 0 children $\iff$ $\variable{treeValues}[i].\variable{predicate} \in
    \predicates{}$;
  \item has 1 child $\iff$ $\variable{treeValues}[i].\variable{predicate} = \neg$;
  \item has $>1$ child $\iff$ $\variable{treeValues}[i].\variable{predicate} \in
    \{ \land, \lor \}$.
  \end{itemize}
\end{constraint}

\begin{constraint}
  For $i = 0, \dots, \maxNumNodes{} - 1$,
  \[
    \variable{treeStructure}[i] \ne i \implies
    \variable{treeValues}[i].\variable{predicate} \ne \top.
  \]
\end{constraint}

If the clause should be disabled, $\variable{numNodes} = 1$ and
$\variable{treeValues}[0].\variable{predicate} = \top$.

\begin{constraint}
  Adjacency matrix representation:
  \begin{align*}
    A[i][j] = 0 \iff &\nexists k: \variable{headsOfClauses}[k].\variable{predicate} = j \text{ and } \\
    &i \in \{ a.\variable{predicate} \mid a \in \variable{bodiesOfClauses}[k].\variable{treeValues} \}.
  \end{align*}
\end{constraint}

\section{Head of a Clause}

\begin{definition} % TODO: variables are always ordered, constants are not.
  The \emph{head} of a clause is defined by two lists:
  \begin{itemize}
  \item $\variable{predicate} \in \predicates \cup \{ \Box \}$, where $\Box$
    denotes a disabled clause.
  \item \variable{variables} of length $|\variables{}|$ and with domain
    $[0, \maxArity{}]$: how many times each variable appears in the head atom.
  \item \variable{constants} of length $\maxArity{}$ and with domain
    $\constants{} \cup \{ \Box \}$, where $\Box$ denotes that the position is
    reserved for a variable.
  \end{itemize}
  We also define the predicate's \variable{arity} using the same
  \variable{table} constraint.
\end{definition}

\begin{constraint}
  For each variable $v \in \variable{variables}$, $v \le \variable{arity}$.
\end{constraint}

\begin{constraint}
  For $i = 0, \dots, \maxArity{} - 1$,
  \[
    i \ge \variable{arity} \implies \variable{constants}[i] = 0.
  \]
\end{constraint}

\begin{constraint}[Connecting the two lists]
  $\sum_{v \in \variable{variables}} v = \variable{count}(\Box,
  \variable{constants}) + \variable{arity} - \maxArity{}$.
\end{constraint}

In \variable{variables}, all zeros must go after all non-zeros. For example,
if we have to pick one variable out of two, we must pick the first one.
\begin{constraint}
  For $i = 0, \dots, |\variables{}| - 2$, and $j = i + 1, \dots,
  |\variables{}| - 1$,
  \[
    \variable{variables}[i] \ne 0 \quad \text{or} \quad \variable{variables}[j]
    = 0.
  \]
\end{constraint}

@@

% Finish the propagation algorithm for conditional independence. The
%  propagation algorithm for independence is extended with masks that are either
%  potential or definite. Masking happens in two stages: first, we mask
%  expressions within formulas, and then predicates. Masking algorithm uses an
%  algorithm for perfect bipartite matching.
